Gravitational field calculation in spherical coordinates using variable densities in depth

### 2018-04-05 manuscript/manuscript.tex
Added Algorithm summary in manuscript

\begin{summary}
We present a new methodology to compute the gravitational fields generated by 
any tesseroid whose density varies continuously in depth, i.e. a forward 
gravity model in spherical coordinates for varying densities on the radial 
direction.
It's based on the numerical approximation of the gravitational fields by 
applying the Gauss-Legendre Quadrature along with a preexisting adaptive 
discretization algorithm and a new density-based discretization algorithm that 
control both the accuracy and the computation time of the forward model.
The first algorithm is regulated by a predefined scalar called distance-size 
ratio ($D$) and reduces the errors due to the distance between the tesseroid 
and the computation point.
While the second one, controlled by a delta ratio ($\delta$), decreases the 
errors introduced by the variation of the density function.
We have also obtained analytical solutions for a spherical shell with variable 
density and compared them with the results of the numerical model in case of a 
linear and an exponential density function.
These comparisons allowed to obtain default values for the distance-size and 
the delta ratio in order to guarantee the accuracy of the forward model.
The resulting optimal values of $D$ for the gravity potential, the components 
of its gradient and the Marussi tensor components are 1, 2 and 8, 
respectively.
While a $\delta=0.2$ is needed in the computation of every gravitational 
field, guaranteeing the accuracy even for high varying exponential densities.
We have finally tested the new software by modelling the Neuqu\'en sedimentary 
basin, located to the east of the Andes between 32$^\circ$S and 40$^\circ$S 
latitude, assigning an exponential density variation to it and computing all 
the gravity fields it generates.
\end{summary}

@@

The main issue that should be overcome is to take into account the curvature 
of the Earth, thus the forward model must be defined in spherical coordinates. 
A common way to achieve this is to discretize the Earth in spherical prisms 
known as tesseroids, which are defined by pairs of latitude, longitude and 
radial boundaries (see Figure \ref{fig:tesseroid-uieda}).
The calculation of the gravity fields generated by an arbitrary tesseroid on 
any external point involves the resolution of volume integrals that are 
generally approximated by numerical computations.
The literature offers two approaches: one involves Taylor series expansion 
\citep{Heck2007, Grombein2013} while the other makes use of Gauss-Legendre 
Quadrature (GLQ) \citep{Asgharzadeh2007, Uieda2016, Uieda2017}.

@@

The spherical prisms known as tesseroids are mass elements defined in a 
geocentric spherical coordinate system bounded by a pair of parallels, 
a pair of  meridians, and two concentric spherical surfaces.
We define an external computation point $P(r, \phi, \lambda)$ where the 
gravity fields generated by the tesseroid are going to be calculated 
with respect to the local north oriented Cartesian coordinate system 
located at $P$.
In the special case of homogeneous density, the gravity potential, its 
gradient and the Marussi tensor components can be calculated through 
the integrals obtained by \citet{Grombein2013} \citep[for same notation 
as the one we will use, see][]{Uieda2016}.

In our case, we will assume that the tesseroid has an arbitrary 
variable density in depth, i.e. only depends on the radial spherical 
coordinate.
Thus, the integrals for the gravity fields are slightly modified:

@@

From equation \ref{eq:glq-var-dens} we can observe that the GLQ 
approximates the gravity fields of a tesseroid as the ones generated by 
$N_\lambda N_\phi N_r$ point masses located at the nodes of the 
Legendre polynomials, where $N_i$ are the orders of the quadrature for 
each integration, with $i \in \{ \lambda, \phi, r \}$, in the same way 
as the GLQ works in the homogeneous density case.
Nevertheless, in this new variable density scenario, the information of 
the density function is summarised into the weights by taking into 
account only the values it assumes on those nodes.
Although this may sound discouraging, we intent to prove otherwise and 
show that this method, along with well suited discretization 
algorithms, approximates the gravity fields with good accuracy.\todo{no 
se si poner esta ultima oracion}

@@

\noindent where $i \in \{\lambda, \phi, r\}$, $d$ is the distance 
between the geometric centre of the tesseroid and the computation 
point, and $D$ is a predefined value called distance-size ratio.
If the inequality \ref{eq:distance-size-ratio} is not held for at least 
one spherical coordinate, then the tesseroid must be divided by half in 
that direction.
On the contrary, if the inequality is satisfied for every spherical 
coordinate, the algorithm computes the desired gravity field using the 
GLQ approximation.
A complete description of the modified adaptive discretization 
algorithm can be found in \citet{Uieda2016}.

@@

If on the contrary the inequation \ref{eq:delta-density} is not held, 
then the tesseroid will not be subdivided.
The same happens if the tesseroid radial dimension is bellow the 
threshold of 1mm. On the first case due to the fact that no more 
discretizations are needed, and on the second one because smaller 
tesseroids will not introduce meaningful increases in the accuracy of 
the computation.
This two conditions guarantee that the algorithm is finite.
Once they are met, every subdividing tesseroid is subjected to the 
modified adaptive discretization algorithm and the GLQ in order to 
compute the gravity field generated by it.

@@

@@

When computing any gravity field generated by an arbitrary tesseroid on a specific computation point, the new software:

\begin{enumerate}
    \renewcommand{\theenumi}{(\arabic{enumi})}
    \item Applies the density-based discretization algorithm in case of variable density producing subdivisions of the original tesseroid,
    \item Applies the modified adaptive discretization algorithm for each subdivision, generating a set of smaller tesseroids,
    \item Computes the desired gravity field for every small tesseroid through the GLQ approximation (equation \ref{eq:glq-var-dens}).
\end{enumerate}

These steps are applied for each computation point, adding the contribution of each tesseroid to the resulting gravity field.

>>>>

\begin{summary}
We present a new methodology to compute the gravitational fields generated by 
any tesseroid whose density varies continuously in depth, i.e. a forward 
gravity model in spherical coordinates for varying densities on the radial 
direction.
It's based on the numerical approximation of the gravitational fields by 
applying the Gauss-Legendre Quadrature along with a preexisting adaptive 
discretization algorithm and a new density-based discretization algorithm that 
control both the accuracy and the computation time of the forward model.
The first algorithm is regulated by a predefined scalar called distance-size 
ratio ($D$) and reduces the errors due to the distance between the tesseroid 
and the computation point.
While the second one, controlled by a delta ratio ($\delta$), decreases the 
errors introduced by the variation of the density function.
We have also obtained analytical solutions for a spherical shell with variable 
density and compared them with the results of the numerical model in case of a 
linear and an exponential density function.
These comparisons allowed to obtain default values for the distance-size and 
the delta ratio in order to guarantee the accuracy of the forward model.
The resulting optimal values of $D$ for the gravity potential, the components 
of its gradient and the Marussi tensor components are 1, 2 and 8, 
respectively.
While a $\delta=0.2$ is needed in the computation of every gravitational 
field, guaranteeing the accuracy even for high varying exponential densities.
We have finally tested the new software by modelling the Neuqu\'en sedimentary 
basin, located to the east of the Andes between 32$^\circ$S and 40$^\circ$S 
latitude, assigning an exponential density variation to it and computing all 
the gravitational fields it generates.
\end{summary}

@@

The main issue that should be overcome is to take into account the curvature 
of the Earth, thus the forward model must be defined in spherical coordinates. 
A common way to achieve this is to discretize the Earth in spherical prisms 
known as tesseroids, which are defined by pairs of latitude, longitude and 
radial boundaries (see Figure \ref{fig:tesseroid-uieda}).
The calculation of the gravitational fields generated by an arbitrary 
tesseroid on any external point involves the resolution of volume 
integrals that are generally approximated by numerical computations.
The literature offers two approaches: one involves Taylor series expansion 
\citep{Heck2007, Grombein2013} while the other makes use of Gauss-Legendre 
Quadrature (GLQ) \citep{Asgharzadeh2007, Uieda2016, Uieda2017}.

@@

The spherical prisms known as tesseroids are mass elements defined in a 
geocentric spherical coordinate system bounded by a pair of parallels, 
a pair of  meridians, and two concentric spherical surfaces.
We define an external computation point $P(r, \phi, \lambda)$ where the 
gravitational fields generated by the tesseroid are going to be calculated 
with respect to the local north oriented Cartesian coordinate system 
located at $P$.
In the special case of homogeneous density, the gravity potential, its 
gradient and the Marussi tensor components can be calculated through 
the integrals obtained by \citet{Grombein2013} \citep[for same notation 
as the one we will use, see][]{Uieda2016}.

In our case, we will assume that the tesseroid has an arbitrary 
variable density in depth, i.e. only depends on the radial spherical 
coordinate.
Thus, the integrals for the gravitational fields are slightly modified:

@@

From equation \ref{eq:glq-var-dens} we can observe that the GLQ 
approximates the gravitational fields of a tesseroid as the ones 
generated by $N_\lambda N_\phi N_r$ point masses located at the nodes 
of the Legendre polynomials, where $N_i$ are the orders of the 
quadrature for each integration, with $i \in \{ \lambda, \phi, r \}$, 
in the same way as the GLQ works in the homogeneous density case.
Nevertheless, in this new variable density scenario, the information of 
the density function is summarised into the weights by taking into 
account only the values it assumes on those nodes.
Although this may sound discouraging, we intent to prove otherwise and 
show that this method, along with well suited discretization 
algorithms, approximates the gravitational fields with good accuracy.
\todo{no se si poner esta ultima oracion}

@@

\noindent where $i \in \{\lambda, \phi, r\}$, $d$ is the distance 
between the geometric centre of the tesseroid and the computation 
point, and $D$ is a predefined value called distance-size ratio.
If the inequality \ref{eq:distance-size-ratio} is not held for at least 
one spherical coordinate, then the tesseroid must be divided by half in 
that direction.
On the contrary, if the inequality is satisfied for every spherical 
coordinate, the algorithm computes the desired gravitational field using the 
GLQ approximation.
A complete description of the modified adaptive discretization 
algorithm can be found in \citet{Uieda2016}.

@@

If on the contrary the inequation \ref{eq:delta-density} is not held, 
then the tesseroid will not be subdivided.
The same happens if the tesseroid radial dimension is bellow the 
threshold of 1mm. On the first case due to the fact that no more 
discretizations are needed, and on the second one because smaller 
tesseroids will not introduce meaningful increases in the accuracy of 
the computation.
This two conditions guarantee that the algorithm is finite.
Once they are met, every subdividing tesseroid is subjected to the 
modified adaptive discretization algorithm and the GLQ in order to 
compute the gravitational field generated by it.

@@

When computing any gravitational field generated by an arbitrary tesseroid on 
a specific computation point, we present the following methodology:

\begin{enumerate}
    \renewcommand{\theenumi}{(\arabic{enumi})}
    \item Apply the density-based discretization algorithm in case of 
          variable density producing subdivisions of the original tesseroid,
    \item Apply the modified adaptive discretization algorithm for 
          each subdivision, generating a set of smaller tesseroids,
    \item Compute the desired gravitational field for every small tesseroid 
          through the GLQ approximation with fixed quadrature orders
          (equation \ref{eq:glq-var-dens}).
\end{enumerate}

In most practical cases, the forward model will consist in a set of 
tesseroids whose gravitational effect will be computed on a set of 
computation points.
These steps can be applied on every tesseroid and for each computation point,
adding the contribution of each tesseroid to the resulting 
gravitational field.

@@
